---
title: 'STAT 216: Introduction to Statistics'
author: "Syllabus for Online Sections: Fall 2020"
output:
  html_document:
      theme: cerulean
      toc: TRUE
      toc_depth: 3
      toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)

library(RColorBrewer)
library(tidyverse)

pal <- brewer.pal(6, 'Pastel1')
```

---

> **_NOTE_**: This syllabus is for online sections of STAT 216 only. For blended sections of STAT 216, refer to the [blended section syllabus](https://mtstateintrostats.github.io/Syllabus/).

---


# Instructor contact information  {.tabset .tabset-fade}
Your primary contact in STAT 216 is your instructor. If you have
concerns that cannot be answered by your instructor, you may reach
out to the Student Success Coordinator or the Course Supervisor.

Refer to your section's _D2L Announcements_ page for your
instructor and co-instructor/TA contact information.

## Student Success Coordinator
[**Jade Schmidt**](http://www.math.montana.edu/directory/faculty/1524571/jade-schmidt){target="_blank"}  
email: jade.schmidt2@montana.edu  
Office: Wilson 2-263  
Phone: (406) 994-6557

---

## Assistant Coordinator
[**Melinda Yager**](http://www.math.montana.edu/directory/faculty/1582830/melinda-yager){target="_blank"}  
email: melinda.yager@montana.edu  
Office: Wilson 1-145  
Phone: (406) 994-6876

---

## Course Supervisor
[**Dr. Stacey Hancock**](http://www.math.montana.edu/shancock/){target="_blank"}  
email: stacey.hancock@montana.edu  
Office: Wilson 2-195  
Phone: (406) 993-5350  


---

# Course calendars

* STAT 216 calendar for students in online sections
* [MSU academic calendar](http://catalog.montana.edu/academiccalendar/)

---

# Course description and objectives
Stat 216 is designed to engage you in the statistical investigation process from developing a research question and data collection methods to analyzing and communicating results. This course introduces basic descriptive and inferential statistics using both traditional (normal and t-distribution) and simulation approaches including confidence intervals and hypothesis testing on means (one-sample, two-sample, paired), proportions (one-sample, two-sample), regression and correlation. You will be exposed to numerous examples of real-world applications of statistics that are designed to help you develop a conceptual understanding of statistics. After taking this course, you should be able to:

- Understand and appreciate how statistics affects your daily life and the fundamental role of statistics in all disciplines
- Evaluate statistics and statistical studies you encounter in your other courses
- Critically read news stories based on statistical studies as an informed consumer of data
- Assess the role of randomness and variability in different contexts
- Use basic methods to conduct and analyze statistical studies
- Evaluate and communicate answers to the four pillars of statistical inference: How strong is the evidence of an effect? What is the size of the effect? How broadly do the conclusions apply? Can we say what caused the observed difference?

### MUS STAT 216 learning outcomes
1. Understand how to describe the characteristics of a distribution.
2. Understand how data can be collected, and how data collection dictates the choice of statistical method and appropriate statistical inference.
3. Interpret and communicate the outcomes of estimation and hypothesis tests in the context of a problem.
4. To understand the scope of inference for a given dataset.

### CORE 2.0
This course fulfills the Quantitative Reasoning (Q) CORE 2.0 requirement because learning probability and statistics allows us to disentangle what's really happening in nature from “noise” inherent in data collection. It allows us to evaluate claims from advertisements and results of polls and builds critical thinking skills which form the basis of statistical inference. Students completing a Core 2.0 Quantitative Reasoning (Q) course should demonstrate an ability to:

Interpret and draw inferences from mathematical models such as formulas, graphs, diagrams or tables.
Represent mathematical information numerically, symbolically and visually.
Employ quantitative methods in symbolic systems such as, arithmetic, algebra, or geometry to solve problems.

 <a href="#top">Back to top</a>
 
 ---
 
# Prerequisites
Entrance to STAT 216 requires at least one of the following be met:

- Grade of C- or better in a 100-level math course (or equivalent)
- Grade of B or better in MATH 096
- Level 30 on the [Math Placement Exam](http://www.montana.edu/testing/MPLEX.html) or a combination of a good score on Math portion of SAT (540 or higher) or ACT (23 or higher) and/or good high school GPA
  + See the [Math Prerequisite Flowchart](http://www.math.montana.edu/undergrad/documents/MHiearchyFlowchart.pdf) for more details.

You should have familiarity with computers and technology (e.g., Internet browsing, word processing, opening/saving files, converting files to PDF format, sending and receiving e-mail, etc.).


 <a href="#top">Back to top</a>
 
 ---
 
# Course materials and resources

### Required Textbook
The required textbook for the course is *Introduction to Statistical Investigations for Montana State University* by Tintle, Chance, Cobb, Rossman, Roy, Swanson, and VanderStoep (Wiley, 2016). MSU negotiated a reduced price for the textbook available only through the MSU bookstore. You have the option to either purchase the textbook from the MSU bookstore or you can purchase an access code for [WileyPlus](https://education.wiley.com/ngonboard/index.html#/Login) (which will contain the e-text and additional resources such as practice problems and videos created by the authors). Other materials, such as readings and videos will be available in D2L. 

**Note that blended sections will be using a downloadable textbook.** Students enrolled in a blended section will need to purchase the activity workbook from the MSU Bookstore, but should not purchase the text by Tintle, et al. See your section's D2L page for more details.

### STAT 216 YouTube channel
Visit our [YouTube channel](https://www.youtube.com/channel/UClGAq0KVt2d2sKd4aIAI4qQ/videos) for short instructional videos on course material (also linked in D2L).

### Learning management tools

- [**Course Webpage**](http://www.math.montana.edu/courses/s216/): Find web links for activities not included in your textbook, data sets used in the course, plus additional resources.  Note the course webpage contains information for both blended and online sections.  Be sure to note the requirements for online students.

- [**D2L**](https://ecat1.montana.edu/): Find your instructor contact info, announcements, RATs, exploration information, instructor notes, exam review material, assignment and data files, discussion forums, gradebook.
  + *Important*: Make sure you are receiving email notifications for any D2L activity. In D2L, click on your name, then Notifications. Check that D2L is using an email address that you regularly check; you have the option of registering a mobile number. Check the boxes to get notifications for announcements, content, discussions, and grades. 
  + If you have a question about the course materials, computing, or logistics, please post your question to your D2L discussion board instead of emailing your instructor. This ensures all students can benefit from the responses. Other students are encouraged to respond. 

- [**Gradescope**](https://www.gradescope.com/): Submit and review assignments, review exam grades. For more details, see our [Gradescope Help for Students](http://www.math.montana.edu/courses/s216/GradescopeHelpForStudents.pdf) document

- [**Web Applets**](http://www.rossmanchance.com/ISIapplets.html): Used on nearly every activity for exploratory and inferential data analysis.
  + Students in a blended section will utilize the statistical software package `R` for exploratory and inferential data analysis.  `R` will also be used in Stat 217.
  
- [**Math Learning Center**](http://www.math.montana.edu/undergrad/mlc/index.html): Free drop-in tutoring for 100- and 200-level math and stat courses.

- [**WileyPlus**](https://education.wiley.com/ngonboard/index.html#/Login): Additional resources for the textbook.  This is a separate fee (unless you purchase the e-text) and is optional.
  
 <a href="#top">Back to top</a>
 
 ---
 
# Course format and organization
 
This is an online course. This perhaps obviously means we will never meet in person. That greatly changes your role as a student as well as mine as an instructor. 

### Weekly Expectations

Each week, online students will be expected to:

* **read** assigned sections of the textbook and **watch videos** on that week's content,
* complete 1 - 2 **RATs** (readiness assessment tests) in [D2L](https://ecat1.montana.edu/) to gauge your understanding of the readings and videos,
* complete 1 - 2 **explorations** in the textbook or provided on [D2L](https://ecat1.montana.edu/),
* complete 1 **discussion post** based on the week's content or a relevant case study,
* complete 1 **assignment** in [Gradescope](https://www.gradescope.com/).

### Online Instructor's Role

Most of the communication in the course should be done using the D2L Discussion and Communication tools. Personal questions you may have can be emailed to your instructor using the D2L Email link. Your instructor will check email periodically, and try to respond to emails as promptly as possible if the email is received during usual daytime or early evening hours. Office hours will be the best way for live-streaming Q&A.

Questions on course material can be posted in the appropriate Discussion folders for that topic. Responses from your instructor to those questions will be posted as a response to that Discussion posting. You are encouraged to post questions on course material in the Discussion folder so that everyone may benefit from the responses! You can post questions and respond to each other's questions. It honestly helps to try to explain things to each other instead of waiting for the instructor’s responses, as trying to explain things out loud helps us learn even if the answer is not correct! We all have a voice in an online learning environment! If you wish to communicate with peers within D2L to discuss homework questions and the like, any messages you send privately to each other are not seen by course instructors. You may even view the Classlist and email other members of the course to ask questions or simply to generally communicate. Only what is posted in Discussion Folders is available for everyone to see. 


### Online Student's Role

It is important for you to understand that your role in this class is not to be a passive note taker because there is no lecture in which to take notes! Each of you is expected to read and understand the material in the text, work on the homework problems, complete all assigned work, and ask questions about concepts you do not understand (and hopefully answer a few questions about the concepts you do understand) in the discussion section of the course. Simply reading the words in the text without comprehension, guessing your way through the assigned work, or looking at the solutions to the homework and thinking you understand how to do the problem is a waste of time and a sure way to fail the course. Your instructor will answer your questions but you have to learn the material on your own. In this class, more than any other course you have taken, you are responsible for your education. 

*Note*: The proper way to learn material presented in this course is to read the chapter, fill out the reading guide, watch the videos, work through the explorations, and then complete all assigned work on your own. Don’t rely only on the feedback to learn the material. Learning in this way will not train you to think independently, and may show you how to do one particular problem, but you will not be able to read another problem and be able to solve it. Point being: statistics is about understanding all aspects of an idea presented, not just a systematic way of using a formula. In other words, this is NOT a math class!

### Group Expectations

Our use of groups in this course is based on educational research which provides strong evidence that working in groups is effective and helps us learn. Additionally, you will build team skills that employers look for when hiring. By expressing your opinions and catching each other’s mistakes, you will learn to communicate statistical concepts. These are partly “common sense” ideas (for instance, gathering more data provides a better foundation for decision making), but they are often phrased in odd ways. We find it really helps to talk about them with others. 

 <a href="#top">Back to top</a>
 
 ---
 
# Course assessment

Your grade in STAT 216 will contain the following components.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
grades <- tribble(
  ~category,  ~percent,
  "D2L discussion", 10,
  "Readiness Assessment Tests (RATs)", 5,
  "Explorations", 10,
  "Assignments", 30,
  "Midterm exams (2)", 25,
  "Final group project/Exam", 20
)
grades$category <- fct_relevel(grades$category, "D2L discussion", "Readiness Assessment Tests (RATs)", "Explorations", "Assignments", "Midterm exams (2)", "Final group project/Exam")
grades$location <- rep(1, 6)
grades %>% ggplot(aes(x=location, y=percent)) +
  geom_col(aes(fill=category), col='black') + 
  guides(fill=FALSE) + 
  geom_text(aes(x=rep(1, 6), 
                label=c("D2L discussion (10%)",
                        "Readiness Assessment Tests (RATs) (5%)",
                        "Explorations (10%)",
                        "Assignments (30%)",
                        "Midterm exams (25%)",
                        "Final group project/Exam (20%)"), 
                y=c(95,87.5,80,60,32.5,10))) + 
  labs(x='', y='Cumulative percent') + 
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(values=pal)
```

1. D2L discussion (10%): Each week, you will either ask or answer questions on a reading, video, exploration, or provided relevant case study, posting questions, answers and discussion in the D2L Discussion Forum.  Discussion posts will be due each **Thursday at 11pm MST**.
2. Readiness Assessment Tests (RATs) (5%): You will be expected to complete the assigned textbook reading and videos prior to completing the exploration(s) for each week. RATs will be taken in [D2L](https://ecat1.montana.edu/) to check your understanding of the material.  *A score of at least 50\% on a RAT will be converted to a 100\% after the due date by your instructor.* RATs for the week will be due each ***Wednesday at 11pm MST.**
3. Explorations (10%): Explorations will primarily be in your textbook, though your instructor will provide some in [D2L](https://ecat1.montana.edu/) also.  Explorations will be due in Gradescope each **Thursday at 11pm MST**.
4. Assignments (30%): You will complete weekly assignments in [Gradescope](https://www.gradescope.com/).  Assignments will be due each **Monday at 11pm MST**.
5. Midterm exams (30%): There will be two midterm exams (worth 12.5% each). **Midterm exam 1 will be during the week of September 22-25; Midterm exam 2 during November 4-10.** Midterm exams will be taken in [Gradescope](https://www.gradescope.com/).  You will have 2 hours from the time you open the exam to complete it.  **Exams must be taken on the scheduled date.  You are responsible for arranging your schedule to accommodating taking of exams.** 
  + The midterm exams are intended as an opportunity for you to consolidate your knowledge from the first and second thirds of our course, so you are highly encouraged to take the exam using only your personal study materials. However, we rarely do statistics in a vacuum, so you will be allowed access to other students, the MLC, and your instructor for help with questions. You may use other online resources, but beware – Google may not be able to help you. Blindly applying formulas you found through a search engine without understanding of the underlying statistical concepts will not earn you full points on an exam question. You are much better off reviewing the course resources posted on D2L.
6. Final group project/Exam (20%): For your final, you will either complete a group project, take a final exam, or some combination of the two.  The exact format will be determined at a later date and this syllabus updated to reflect the decision.  Your instructor will e-mail you and post the format of the final exam to D2L at least 5 weeks prior to the last day of 'class'.  

*Plan ahead*: In our experience, it takes 6 to 9 hours per week outside of class to achieve a good grade in Stat 216. As this is an online course, add the typical 'in class' time of 3 hours per week to that estimate.  By “good” we mean at least a C because a grade of D or below does not count toward fulfilling degree requirements. Many of you set your goals higher than just getting a C, and we fully support that. You need roughly 9 to 12 hours per week to review past activities, read feedback on previous assignments, complete current assignments, and prepare for the next week's class. 

### Late work policy

In general, make-up exams or late homework assignments will not be allowed. Case-by-case exceptions may be granted in only extreme cases at the discretion of the instructor (daily work) or Student Success coordinator (exams). You must speak to your instructor or Student Success coordinator on or before the due date or exam date and provide documentation explaining your absence for the instructor or Student Success coordinator to determine whether an exception should be granted. If you fail to provide documentation as requested then you will not be able to make-up missed work at all.   

### Letter grades

Final course grades will be determined according to the following scale.

Letter Grade|Weighted Score
---|---
A|93-100%
A-|90-92.99%
B+|87-89.99%
B|83-86.99%
B-|80-82.99%
C+|77-79.99%
C|73-76.99%
C-|70-72.99%
D|60-69.99%
F|<59.99%

The grade cutoffs may be shifted _downward_ at the end of the semester based on student performance (never upward).


<a href="#top">Back to top</a>
 
---
 

# Diversity and inclusivity statements

Respect for Diversity: It is our intent that students from all diverse backgrounds and perspectives be well-served by this course, that students' learning needs be addressed both in and out of class, and that the diversity that students bring to this class be viewed as a resource, strength and benefit. It is our intent to present materials and activities that are respectful of diversity: gender identity, sexual orientation, disability, age, socioeconomic status, ethnicity, race, religion, culture, perspective, and other background characteristics. Your suggestions about how to improve the value of diversity in this course are encouraged and appreciated. Please let us know ways to improve the effectiveness of the course for you personally or for other students or student groups.

In addition, in scheduling exams, we have attempted to avoid conflicts with major religious holidays. If, however, we have inadvertently scheduled an exam or major deadline that creates a conflict with your religious observances, please let us know as soon as possible so that we can make other arrangements. 

Support for Inclusivity: We support an inclusive learning environment where diversity and individual differences are understood, respected, appreciated, and recognized as a source of strength. We expect that students, faculty, administrators and staff at MSU will respect differences and demonstrate diligence in understanding how other peoples' perspectives, behaviors, and worldviews may be different from their own.

<a href="#top">Back to top</a>
 
---

# Policy on academic misconduct

Students in an academic setting are responsible for approaching all assignments with rigor, integrity, and in compliance with the University Code of Student Conduct. This responsibility includes:

1. consulting and analyzing sources that are relevant to the topic of inquiry;
2. clearly acknowledging when they draw from the ideas or the phrasing of those sources in their own writing;
3. learning and using appropriate citation conventions within the field in which they are studying; and
4. asking their instructor for guidance when they are uncertain of how to acknowledge the contributions of others in their thinking and writing.

When students fail to adhere to these responsibilities, they may intentionally or unintentionally "use someone else’s language, ideas, or other original (not common-knowledge) material without properly acknowledging its source" [http://www.wpacouncil.org](http://www.wpacouncil.org). When the act is intentional, the student has engaged in plagiarism.

Plagiarism is an act of academic misconduct, which carries with it consequences including, but not limited to, receiving a course grade of “F” and a report to the Office of the Dean of Students. Unfortunately, it is not always clear if the misuse of sources is intentional or unintentional, which means that you may be accused of plagiarism even if you do not intentionally plagiarize. If you have any questions regarding use and citation of sources in your academic writing, you are responsible for consulting with your instructor before the assignment due date. In addition, you can work with an MSU Writing Center tutor at any point in your writing process, including when you are integrating or citing sources. You can make an appointment and find citation resources at www.montana.edu/writingcenter.

**In STAT 216, students involved in plagiarism on assignments (all parties involved) will receive a warning on the first offense and a 50% reduction in score. The second offense will result in a zero on that assignment, and the incident will be reported to the Dean of Students. Academic misconduct on an exam will result in a zero on that exam and will be reported to the Dean of Students, without exception.**

[More information about Academic Misconduct from the Dean of Students](https://www.montana.edu/deanofstudents/syllabus.html)

<a href="#top">Back to top</a>
